---
title: "Crab Age Prediction Appendix: R Markdown"
author: "200018630 - Group 4"
date: "March 30, 2024"
output: pdf_document
---

## Introduction

This document presents an analysis for predicting the age of crabs using various statistical techniques and data visualization methods. The analysis involves data preprocessing, exploratory data analysis (EDA), dimensionality reduction, clustering, and decision tree model.

### Library Import

Import the necessary libraries. 

```{r setup, include=TRUE}
library(dplyr)
library(GGally)
library(aplpack)
library(corrplot)
library(rpart)
library(rpart.plot)
library(factoextra)
```

### Data Preparation

Load data, set directory, and preprocess:

```{r data-preparation, include=TRUE}
folder_path <- file.path(dirname(rstudioapi::getSourceEditorContext()$path), "/Group Project ")
setwd(folder_path)

# Load and preprocess data
Crab_Data <- read.csv('CrabAgePrediction.csv') 
Crab_Data <- Crab_Data %>%
  mutate(Sex = as.factor(Sex)) %>%  # binary column
  mutate(Female = as.integer(Sex == 'F'),  
         Male = as.integer(Sex == 'M'),   
         Indeterminate = as.integer(Sex == 'I'))
```

## Exploratory Data Analysis (EDA)

### Correlation Matrix

Correlation matrix of the numerical columns.

```{r correlation-matrix, include=TRUE}
Crab_Data_Corr <- Crab_Data[, sapply(Crab_Data, is.numeric)]
corrplot(cor(Crab_Data_Corr), method = 'circle')
```

### Scatterplot Matrix

Scatterplot matrix to understand pairwise relationship between variables:

```{r scatterplot-matrix, include=TRUE}
set.seed(123)
sample_size <- 30 # Sample
Crab_Data_Sample <- Crab_Data_Corr[sample(1:nrow(Crab_Data), sample_size), ]
pairs(Crab_Data_Sample)
```

### Parallel Coordinates & Chernoff Faces

Parallel Coordinates and Chernoff Faces are used for multivariate data visualization:

```{r parallel-coordinates-chernoff-faces, include=TRUE}
Crab_Data_Sample_Norm <- as.data.frame(scale(Crab_Data_Sample))

ggparcoord(data = Crab_Data_Sample_Norm %>% select_if(is.numeric), 
           groupColumn = NULL) 
faces(Crab_Data_Sample_Norm)
```

### Multidimensional Scaling (MDS) and PCA

Multidimensional Scaling (MDS) and Principal Component Analysis (PCA) to visualize the data in lower dimensions:

```{r mds-pca, include=TRUE}
# MDS
dist_matrix <- dist(Crab_Data_Sample_Norm)  # Distance matrix
mds_result <- cmdscale(dist_matrix)  # Apply MDS
plot(mds_result[,1], mds_result[,2], xlab="Dimension 1", ylab="Dimension 2", 
     main="Sample MDS Plot", type="n")
text(mds_result[,1], mds_result[,2])

Crab_Data_PCA_Sample <- prcomp(Crab_Data_Sample, scale. = TRUE)
pca_scores <- Crab_Data_PCA_Sample$x[, 1:2] 
plot(pca_scores, main = 'Sample PCA Biplot', type="n")  
text(pca_scores[,1], pca_scores[,2])
```

## Dimensionality Reduction and Clustering

Perform PCA on the full dataset and explore clustering techniques to identify data patterns:

```{r dimensionality-reduction-clustering, include=TRUE}
# PCA on Full Data
Crab_Data_PCA <- prcomp(Crab_Data_Corr, scale. = TRUE)
biplot(Crab_Data_PCA, main = 'PCA Biplot')

# Proportion of variance explained by each PCs
variance_explained <- Crab_Data_PCA$sdev^2 / sum(Crab_Data_PCA$sdev^2)
cumulative_variance <- cumsum(variance_explained)
# Cumulative variance plot
plot(cumulative_variance, xlab = "Number of Principal Components", 
     ylab = "Cumulative Variance Explained",
     type = "b", pch = 19, main = "Cumulative Variance Explained by PCA Components")
abline(h = 0.9, col = "red", lty = 2)  # Add a line at 90% variance explained for reference

# K-means Clustering based on PCA
Crab_Data_PCA_Reduced <- Crab_Data_PCA$x[, 1:2] # 2 PC
kmeans_result <- kmeans(Crab_Data_PCA_Reduced, centers = 2)
plot(Crab_Data_PCA_Reduced[, 1], Crab_Data_PCA_Reduced[, 2], col = kmeans_result$cluster, 
     main = "K-means Clustering on PCA Results")
```

### Optimal Clusters Visualization

Use silhouette and elbow methods to determine the optimal number of clusters:

```{r optimal-clusters-visualization, include=TRUE}
# Visualize the average silhouette method
fviz_nbclust(Crab_Data_PCA_Reduced, kmeans, method = "silhouette", k.max = 20) +
  ggtitle("Silhouette Method for Optimal Clusters - PCA")
fviz_nbclust(Crab_Data_Corr, kmeans, method = "silhouette", k.max = 20) +
  ggtitle("Silhouette Method for Optimal Clusters")
# Visualize the within-cluster sum of squares method (elbow method)
fviz_nbclust(Crab_Data_PCA_Reduced, kmeans, method = "wss", k.max = 20) + 
  ggtitle("Elbow Method for Optimal Clusters - PCA")
fviz_nbclust(Crab_Data_Corr, kmeans, method = "wss", k.max = 20) +
  ggtitle("Elbow Method for Optimal Clusters")
```

## Model Building: Decision Tree

Build a decision tree model to predict crab age:

```{r model-building, include=TRUE}
set.seed(123)
train_indices <- sample(1:nrow(Crab_Data), 0.7 * nrow(Crab_Data))
train_data <- Crab_Data[train_indices, ]
test_data <- Crab_Data[-train_indices, ]

model <- rpart(Age ~ ., data = train_data, method = "anova")

# Decision Tree Result
rpart.plot(model, main="Decision Tree for Crab Age Prediction")
predictions <- predict(model, test_data)
rmse <- sqrt(mean((predictions - test_data$Age)^2))
cat("RMSE on test data:", rmse, "\n")
```
